---
title: Document 2- R codes for the asymptotic analyses of a stage-structured matrix
  model for spotted hyenas infected with CDV
author: "S. Benhaiem, L. Marescot"
date: '`r Sys.time()`'
output:
  pdf_document: default
  html_document:
    code_folding: hide
  word_document: default
---

<style>
  /* Style the linenumber div */

  .linenumbers {
    border: 1px solid #ccc;
    border-radius: 4px;
    background-color: #EBEBEB;
    text-align: center;
    padding: 0px 3px;
    font-family: monospace;
    float: left;
    position: absolute;
    transform:translate(-125%);
    font-size: inherit !important;
  }

.spoiler, .spoiler > * { transition: color 0.5s, opacity 0.5s }
.spoiler:not(:hover) { color: transparent }
.spoiler:not(:hover) > * { opacity: 0 }
/* fix weird transitions on Chrome: */
blockquote, blockquote > *:not(a) { color: black }
</style>


This code presents the asymptotic analysis of the matrix model. For each epidemic period, we first load the input files containing the MECMR parameter estimates for that period and then the R script to construct the matrix model.
  
This code is structured as follows:

  a) Checking assumptions (irreducibility and ergodicity) 
  b) Asymptotic analyses (population's growth rate (($\lambda$)), R0, stable stage distribution and reproductive values)
    i) Pre-epidemic
    ii) Epidemic
    iii) Post-epidemic
    iv) Short-term population dynamics (Fig 4)

Note that parameter and submatrix names may differ between main text and R codes.

--------------------------------------------------------------------------------------------------------

```{r, message=FALSE}
library(popdemo)
library(popbio)

source('Input_90-92(pre-epidem).R') 
source('1_Model construction.R') 
M.finalpre<-M.final
NGMpre<-NGM
source('Input_93-94(epidem).R') 
source('1_Model construction.R') 
M.finalepidem<-M.final
NGMepidem<-NGM
source('Input_95-99(post-epidem).R') 
source('1_Model construction.R')
M.finalpost<-M.final
NGMpost<-NGM
```


##a) Checking assumptions (irreducibility and ergodicity)##

Matrix models are termed 'irreducible' when their associated life cycles contain the transition rates to facilitate pathways from all states to all other states. Irreducible matrices are 'ergodic': the stable asymptotic growth rate is independent from the initial stage structure in the population projection. Both conditions should ideally be met for further analyses.

Pre-epidemic period:
Is the matrix model irreducible?
```{r}
isIrreducible(M.finalpre) 
```
Is the matrix model ergodic?
```{r}
isErgodic(M.finalpre)  
```
Epidemic period:
Is the matrix model irreducible?
```{r}
isIrreducible(M.finalepidem) 
```
Is the matrix model ergodic?
```{r}
isErgodic(M.finalepidem)  
```

  
Post-epidemic period:
Is the matrix model irreducible?
```{r}
isIrreducible(M.finalpost) 
```
Is the matrix model ergodic?
```{r}
isErgodic(M.finalpost)  
```

  
##b) Asymptotic analyses (population's growth rate (($\lambda$)), R0, stable stage distribution and reproductive values)## 

For each epidemic period, we calculate 1) the population's growth rate (($\lambda$)), which is the dominant eigenvalue of the final matrix, 2) The basic reproduction number R0, which is the dominant eigenvalue of the next generation matrix, 3) the stable stage distribution, which is the right eigenvector of the final matrix and 4) the reproductive values, which are the left eigenvector of the final matrix.

Note that the values of ($\lambda$) and R0 differ from the ones presented in the main text (Figure 1), as we presented there the mean + SD values estimated via stochastic analyses.

*i) Pre-epidemic* 

($\lambda$):
```{r}
lambda(M.finalpre) 
```

R0:
```{r}
lambda(NGMpre)
```

Stable-stage distribution:
```{r}
rightpre<-as.data.frame(stable.stage(M.finalpre))
rightpre
```

Reproductive values:
```{r}
ReproValuespre <- as.data.frame(reproductive.value(M.finalpre))
ReproValuespre
```

*ii) Epidemic* 

($\lambda$):
```{r}
lambda(M.finalepidem) 
```

R0:
```{r}
lambda(NGMepidem)
```

Stable-stage distribution:
```{r}
rightepidem<-as.data.frame(stable.stage(M.finalepidem))
rightepidem
```

Reproductive values:
```{r}
ReproValuesepidem <- as.data.frame(reproductive.value(M.finalepidem))
ReproValuesepidem
```


*ii) Post-epidemic* 

($\lambda$):
```{r}
lambda(M.finalpost) 
```

R0:
```{r}
lambda(NGMpost)
```

Stable-stage distribution:
```{r}
rightpost<-as.data.frame(stable.stage(M.finalpost))
rightpost
```

Reproductive values:
```{r}
ReproValuespost <- as.data.frame(reproductive.value(M.finalpost))
ReproValuespost
```

```{r , echo= FALSE} 

#here we select the proportion in low ranking states for the first pre-epidemic phase
proLpre<-sum(rightpre[c(1,2,5,6,7,11,12,13,17,18,19),])
# proportion of individuals with low ranking status
proLepidem<-sum(rightepidem[c(1,2,5,6,7,11,12,13,17,18,19),])
proLpost<-sum(rightpost[c(1,2,5,6,7,11,12,13,17,18,19),])


#pre-peidemic period
susc_pre<-sum(rightpre[c(1,3,5,8,11,14,17,20),])  # proportion of susceptible individuals
inf_pre <-sum(rightpre[c(2,4,6,9,12,15,18,21),]) # proportion of infected individuals
recov_pre <-sum(rightpre[c(7,10,13,16,19,22),]) # proportion of recovered individuals
# epidemic period
susc_epidem<-sum(rightepidem[c(1,3,5,8,11,14,17,20),])  # proportion of susceptible individuals
inf_epidem <-sum(rightepidem[c(2,4,6,9,12,15,18,21),]) # proportion of infected individuals
recov_epidem <-sum(rightepidem[c(7,10,13,16,19,22),]) # proportion of recovered individuals
#post-epidemic period
susc_post<-sum(rightpost[c(1,3,5,8,11,14,17,20),])  # proportion of susceptible individuals
inf_post <-sum(rightpost[c(2,4,6,9,12,15,18,21),]) # proportion of infected individuals
recov_post <-sum(rightpost[c(7,10,13,16,19,22),]) # proportion of recovered individuals

#pre-peidemic period
cubs_pre<-sum(rightpre[c(1:4),])  # proportion of cubs 
subadults_pre<-sum(rightpre[c(5:10),]) # proportion of subadults
breeders_pre<-sum(rightpre[c(11:16),]) # proportion of breeders
non.breeders_pre<-sum(rightpre[c(17:22),]) # proportion of non breeders
#peidemic period
cubs_epidem<-sum(rightepidem[c(1:4),])  # proportion of cubs 
subadults_epidem<-sum(rightepidem[c(5:10),]) # proportion of subadults
breeders_epidem<-sum(rightepidem[c(11:16),]) # proportion of breeders
non.breeders_epidem<-sum(rightepidem[c(17:22),]) # proportion of non breeders
#post-peidemic period
cubs_post<-sum(rightpre[c(1:4),])  # proportion of cubs 
subadults_post<-sum(rightepidem[c(5:10),]) # proportion of subadults
breeders_post<-sum(rightepidem[c(11:16),]) # proportion of breeders
non.breeders_post<-sum(rightepidem[c(17:22),]) # proportion of non breeders

#-------------------------------
# reproductive values
reproductive.value<-cbind(reproductive.value(M.finalpre), reproductive.value(M.finalepidem), reproductive.value(M.finalpost))
colnames(reproductive.value)<-c("pre-epidemic", "epidemic", "post-epidemic")
#-------------------------------
```

Summary table for $\lambda$, R0 and stable stage distribution for each epidemic period:


Population indicator       | Pre-epidemic     | Epidemic  | Post-epidemic 
-----------------------    | --------------   | --------  | ----------
$\lambda$                    | `r lambda(M.finalpre)` |`r lambda(M.finalepidem)`|`r lambda(M.finalpost)`
R0                   | `r lambda(NGMpre)` |`r lambda(NGMepidem)`|`r lambda(NGMpost)`
Stable stage:Low-ranking                        | `r proLpre`   |  `r proLepidem`  | `r proLpost` 
Stable stage:High-ranking                       |`r 1-proLpre`  | `r 1-proLepidem`| `r 1-proLpost` 
Stable stage:Susceptible   |  `r susc_pre`   |  `r susc_epidem`| `r susc_post` 
Stable stage:Infected      |   `r inf_pre`   |  `r inf_epidem` |  `r inf_post`
Stable stage:Recovered     |   `r recov_pre`   |  `r recov_epidem` |  `r recov_post`
Stable stage:Cubs  |  `r cubs_pre`   |  `r cubs_epidem`| `r cubs_post` 
Stable stage:Subadults      |   `r subadults_pre`   |  `r subadults_epidem` |  `r subadults_post`
Stable stage:Breeders     |   `r breeders_pre`   |  `r  breeders_epidem` |  `r  breeders_post`
Stable stage:Non-Breeders |   `r non.breeders_pre`   |  `r  non.breeders_epidem` |  `r non.breeders_post`



*iv) Short-term population dynamics* 

```{r}
#Initial values from E-surge's excel file 

initialvalues <-c(0.12639981,0.12639981,# CL
      0.161066754,0.161066754, # CH
      0.002133319,0.002133319,0.002133319,# SAL
      0.00533333, 0.00533333, 0.00533333,# SAH
      0.005333321, 0.005333321, 0.005333321,# BL
      0.007629848, 0.007629848, 0.007629848, # BH
      0.010133336, 0.010133336, 0.010133336,# NBL
      0.012266641, 0.012266641, 0.012266641)# NBH


# Modifying the stage.vector.plot function to enhance figure
stage.vector.plot.mod <-
function (stage.vectors, proportions = TRUE,  
            ylim = NULL, xlab = "Years", ylab = NULL, col = rainbow(8), 
            ...) 
  {
    p <- stage.vectors
    n <- dim(p)[1]
    if (is.null(n)) {
      stop("stage.vectors should be a matrix with two or more stages")
    }
    x <- colnames(p)
    if (is.null(x)) {
      x <- 0:(dim(p)[2] - 1)
    }
    if (length(col) < n) {
      col <- rep(col, n)
    }
    if (proportions) {
      if (is.null(ylab)) {
        ylab <- "Proportion in stage class"
      }
      p <- prop.table(p, 2)
      if (is.null(ylim)) {
        ylim = c(min(p, na.rm = TRUE), max(p, na.rm = TRUE))
      }
      plot(x, p[1, ], type = "n", ylim = ylim, xlab = xlab, 
           ylab = ylab, ...)
    }
    else {
      if (is.null(ylab)) {
        ylab <- "Number in stage class"
      }
      if (is.null(ylim)) {
        ylim = c(floor(min(p, na.rm = TRUE)), ceiling(max(p, 
                                                          na.rm = TRUE)))
      }
      plot(x, p[1, ], type = "n", ylim = ylim, xlab = xlab, 
           ylab = ylab, ...)
    }
    y <- sort(apply(p, 1, mean, na.rm = TRUE), index.return = TRUE, 
              decreasing = TRUE)
    for (i in y$ix) {
      lines(x, p[i, ], lty = 1, col = col[i], lwd = 2)
    }
    leg.names <- paste(names(y$x), "")
    if (leg.names[1] == " ") {
      leg.names <- paste("row", y$ix, "")
    }
    
  }

# The function transdyn is to get the SIR states for each period for the short term projections

transdyn<-function(M.final,n,horizon)
{
  p<-pop.projection(M.final,n,horizon)# Project over 11 years

 all <-as.data.frame(p$stage.vectors)
 colnames(all) <- paste0(c(0:10)) # from 0 to x-1

## We create a new dataframe with 3 rows, S,I,R, which are the sum of the detailed states
# first empty table 

 sir <- matrix(0, ncol = 11, nrow = 3) #col = x
 colnames(sir) <- paste0(c(0:10)) # from 0 to x-1
 rownames(sir) <- c("S","I","R")

#then we fill it
  for (i in names(all)){
  sir[1,i] <- round(sum(all[c(1,3,5,8,11,14,17,20),i]),digits=2) # Susceptible
  sir[2,i] <- round(sum(all[c(2,4,6,9,12,15,18,21),i]),digits=2) # Infected
  sir[3,i] <- round(sum(all[c(7,10,13,16,19,22),i]),digits=2) # Recovered
  }
  return(sir)
}

horizon<-11 # Number of projected years

sirpre<-transdyn(M.finalpre,initialvalues,horizon) 

sirepidem<-transdyn(M.finalepidem,initialvalues,horizon)

sirpost<-transdyn(M.finalpost,initialvalues,horizon)
```

Figure 4 - Short-term dynamics (first ten years) of the proportion of different infection states and their convergence to a stable stage distribution (using the time invariant initial state vector projection from the MECMR model), for susceptible (blue), infected (red) and recovered (green) females (across all demographic and social states) during a) pre-epidem, b) epidem, c) post-epidem.

```{r, echo = FALSE, fig.keep='last'}
#Final figure with 3 panels side by side

nf <- layout(matrix(c(0,0,0,0,1,0,0,2,0,0,3,0,0,0,0),3, byrow = F),widths= c(0.1,1.8,1.8,1.8,0.1), heights=c(1,3.5,1,1,0)) # widths for columns,heights for rows 

# Pre-epidem
par(mar=c(5.1,5.1,4.1,2.1))

stage.vector.plot.mod(sirpre, proportions=T, 
                    cex.lab =1.8,cex.axis=1.8,col=c("blue","red","dark green"),
                    las =1,xlim=c(0,10),ylab="")

mtext(side = 2,text=c("Proportion"), line =4,cex = 1.3)
mtext(side = 3,text=c("a) pre-epidem"), line =1.7,cex = 1.3)

# epidem
par(mar=c(5.1,5.1,4.1,2.1))

stage.vector.plot.mod(sirepidem, proportions=T, 
                    cex.lab =1.8,cex.axis=1.8,col=c("blue","red","dark green"),
                    las =1,xlim=c(0,10),ylab="")

mtext(side = 2,text=c("Proportion"), line =4,cex = 1.3)
mtext(side = 3,text=c("b) epidem"), line =1.7,cex = 1.3)


# post-epidem
par(mar=c(5.1,5.1,4.1,2.1))

stage.vector.plot.mod(sirpost, proportions=T, 
                    cex.lab =1.8,cex.axis=1.8,col=c("blue","red","dark green"),
                    las =1,xlim=c(0,10),ylab="")

mtext(side = 2,text=c("Proportion"), line =4,cex = 1.3)
mtext(side = 3,text=c("c) post-epidem"), line =1.7,cex = 1.3)

```


## References (R packages) ##


Stubben, C. & Milligan, B. Estimating and analyzing demographic models using the popbio package in R. J. Stat. Softw 22, 1-23 (2007).

Stott, I., Hodgson, D. J. & Townley, S. popdemo: an R package for population demography using projection matrix analysis. Methods Ecol. Evol. 3, 797-802 (2012).


















